Cartogram Mapping and its Application to Cancer Data Visualization

### 2019-12-08 version1/version2.Rmd
begin response

---
title: "Cartogram Mapping and its Application to Cancer Data Visualization"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  word_document:
    toc: no
    toc_depth: '3'
link-citations: yes
fontsize: 11pt
csl: vancouver.csl
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(cowplot)
library(png)
library(grid)
```
<!--

Please note that the submission deadline is September 30, 2019.

Author notes:
6000 words maximum.
focused  application  of  literature  to address  a  relevant  clinical  issue:
Mapping cancer outcomes
Cartogram applications should point to examples of cancer outcomes, with the main cartogram section discussing sources considered in grey literature review. Examples of political areas may be considered if cancer data is aggregated at these levels.

- Future steps should consider applications not currently utilized for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualizing those spatial patterns, or the uncertainty in those patterns
- a more applied-focused paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

\section*{Abstract}
<!-- 119 words -->
<!--
Choropleth maps are widely used in cancer atlases to communicate cancer statistics over geographic domains. The type of statistic displayed varies immensely, and effective choices can assist in the understanding of impact of cancer on a population. Atlases have moved from print media to web format, and the availability of graphical user interfaces, and user interaction and animation, greatly enhance the ability to communicate more information and positively engage readers. More use of alternatives to choropleth maps, such as cartograms, could be employed in online atlases, for better representation of spatial distributions, and show statistics on small areas. This is especially true for the geography of Australia, where the population is concentrated is small areas along the coastlines.  
-->

Cancer atlases communicate cancer statistics over geographic domains. These domains are often subdivided by administrative areas such as countries, states or suburbs.
When communicating human-related statistics in Australia, the geographic map base may highlight the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they may be important to consider if they are more densely populated. To

@@

<!-- written below The Atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas, the Standardized Incidence Ratios (SIR) were modeled via Bayesian methods.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas. 
-->

@@

<!-- While there is significant variation across these maps in terms of the measure represented, the scale of the map, the color choices, the legend design and supplementary/supporting statistics, the mapping approach has not changes and almost exclusively only choropleth maps are used. Unlike innovation in spatial statistical methods, interactivity, data collection and record keeping little work has been done to implement different mapping approaches -->

@@

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections - I think this is not relevant here
Choropleths may be considered true topological maps however,, if the landmass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of landmass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection. 
-->

@@

<!-- 
An intentional goal when creating the 1966 Census population cartogram for Canada was to maintain contiguity, while attempting to keep the actual shape of places. The end result was a 'very accurate isodemo-graphic map of Canada'. This intentional design goal coincided with the rising interest in urban geography and presentation of social statistics. Figure \ref{fig:ggcartograms} a shows a population contiguous cartogram of the United States. All states are visible and the shape of the United States overall is still recognizable.
In contrast, Figure \ref{fig:auscartograms} a shows an Australian contiguous cartogram also based on population. The south east is completely distorted, areas with low population are still large very the map, as their initial area was used to determine the area allocated in the contiguous cartogram.

To be able to recognise the significant changes, a reader will usually have to know the initial geography to find the differences in the new cartogram layout [@NAC]. Tobler's Conformal mapping means to preserve angles locally so that the shapes of small areas on a traditional map and a cartogram would be similar.
@CBATCC presents this issue as conflicting tasks or aims, to adjust region sizes and retain region shapes. 
-->

@@

<!-- The shapes are valuable for recognition and allows users to orient themselves on the display. Map creators can efficiently communicate with this kind of map by keeping the outlines or particular elements of the original in the new shape [@NISCC].
The scale of the areas does not impact on the shape recognition.
However it may impact on the visibility of all areas if small areas expand beyond their boundaries. 
Figure \ref{fig:ggcartograms}b shows small overlaps in the North-East region of the United States. These overlaps are also shown in the non-contiguous cartogram of Australia in Figure \ref{fig:auscartograms} b.
-->

@@

A similar method, where each geographic area is represented using a square, tessellated to create a square grid. This method has been used broadly in the media, e.g. FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR.
Each area is represented by a square of the same dimensions.
<!--
Figure \ref{fig:ggcartograms}d shows a hexagon tilemap representing each state using a single hexagon.

Recommended criteria to contrast mapping methods include average cartographic error, and maximum cartographic error, correct adjacency, maximum aspect ratio, and suitable relative positions [@ORC]. However, this does not consider the issues with actually producing rectangular cartograms. Algorithms for the creation of rectangular cartograms
-->

## tilemap
<!-- 173 words -->
A tilemap provides a tessellated display of consistent shapes.
Each tile is usually one unit of measurement, this could be geographic regions such as states, or population-based where one tile is used for a consistent measure of population.
<!-- cite mosaic cartograms -->

A simple tilemap presents the areas in a tessellated grid display, where geographic neighbors are found next to each other, with some necessary displacement employed for regions with more than four neighbors.
These tiles may be labeled or colored to represent a value.
tilemaps may be difficult to create, they are best created manually, with additional time and care required as the number of geographic areas to include increases.
Cano and others [@MDAC] define the term 'mosaic cartograms' for hexagonal tile displays, where the number of tiles for each area can be used to communicate the statistic of regions. 
The complexity of the boundaries can be adjusted in the resulting display, as the size of the tiles used allows a trade-off to be made between boundary complexity and simplicity.

@@

## Geofacet
<!-- 166 words -->
Hafen [@IGF] formalizes the term geofaceting to describe a grid display, the arrangement of tiles to create a grid that mimics the geographic topology of the set of areas. 
Like tilemaps, the arrangement of the areas can be reused for other spatial distribution visualizations.
Geofaceting has the functionality of facets, often used to replicate visualizations for each subset of the data. In this case, the data subsets are geographic.
The amount of information able to be communicated has increased from one value per region in a tilemap to one visualization, this is a more flexible display.
Virtually any plot display can be fit into the tile representing the areas, allowing displays of multiple variables or values per geographic entity.
<!-- As the amount of areas increase care must be taken to ensure legibility of the displays.
@RPIG acknowledges that this display suffers when the areas have very irregular shapes and large size disparities CORRECT reference is Cheng's thesis-->
Manual creation of the grid is required.

@@

<!-- Solutions have been provided for bi-variate displays for movement data, containing Origin - Destination matrices and flow map representations [@MMTVCG]. 
This method makes the maps and matrix connections clear, and easy to link geographic areas with link lines not crossing. Bar charts can also be added to these to show group totals.
-->

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualization"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  word_document:
    toc: no
    toc_depth: '3'
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
link-citations: yes
fontsize: 11pt
csl: vancouver.csl
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(cowplot)
library(png)
library(grid)
```

\section*{Abstract}
<!-- 119 words -->

Cancer atlases communicate cancer statistics over geographic domains. These domains are often subdivided by administrative areas such as countries, states or suburbs.
When communicating human-related statistics in Australia, the geographic map base may highlight the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they may be important to consider if they are more densely populated.

@@

@@

@@

While there is significant variation across these maps in terms of the measure represented, the scale of the map, the color choices, the legend design and supplementary/supporting statistics, the mapping approach has not changes and almost exclusively only choropleth maps are used. Unlike innovation in spatial statistical methods, interactivity, data collection and record keeping little work has been done to implement different mapping approaches.

@@

@@

@@

The shapes are valuable for recognition and allows users to orient themselves on the display. Map creators can efficiently communicate with this kind of map by keeping the outlines or particular elements of the original in the new shape [@NISCC].
The scale of the areas does not impact on the shape recognition.
However it may impact on the visibility of all areas if small areas expand beyond their boundaries. 
Figure \ref{fig:ggcartograms}b shows small overlaps in the North-East region of the United States. These overlaps are also shown in the non-contiguous cartogram of Australia in Figure \ref{fig:auscartograms} b.

@@

A similar method, where each geographic area is represented using a square, tessellated to create a square grid. This method has been used broadly in the media, e.g. FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR.
Each area is represented by a s

## Hexagon tilemap
<!-- 173 words -->
A tilemap provides a tessellated display of consistent shapes.
Each tile is usually one unit of measurement, this could be geographic regions such as states, or population-based where one tile is used for a consistent measure of population.
<!-- cite mosaic cartograms -->

A simple tilemap presents the areas in a tessellated grid display, where geographic neighbors are found next to each other, with some necessary displacement employed for regions with more than four neighbors.
These tiles may be labeled or colored to represent a value.
tilemaps may be difficult to create, they are best created manually, with additional time and care required as the number of geographic areas to include increases.
Cano and others [@MDAC] define the term 'mosaic cartograms' for hexagonal tile displays, where the number of tiles for each area can be used to communicate the statistic of regions. 
The complexity of the boundaries can be adjusted in the resulting display, as the size of the tiles used allows a trade-off to be made between boundary complexity and simplicity.
quare of the same dimensions.

Figure \ref{fig:ggcartograms}d shows a hexagon tilemap representing each state using a single hexagon.

@@

## Geofacet
<!-- 166 words -->
Hafen [@IGF] formalizes the term geofaceting to describe a grid display, the arrangement of tiles to create a grid that mimics the geographic topology of the set of areas. 
Like tilemaps, the arrangement of the areas can be reused for other spatial distribution visualizations.
Geofaceting has the functionality of facets, often used to replicate visualizations for each subset of the data. In this case, the data subsets are geographic.
The amount of information able to be communicated has increased from one value per region in a tilemap to one visualization, this is a more flexible display.
Virtually any plot display can be fit into the tile representing the areas, allowing displays of multiple variables or values per geographic entity.
As the amount of areas increase care must be taken to ensure legibility of the displays.
@IVMVTSAD acknowledges that this display suffers when the areas have very irregular shapes and large size disparities.
Manual creation of the grid is required.

@@
